---
title: "My Start to R"
subtitle: |
  A short introduction to my blog, and R journey. 
date: 01-24-2020

---

Today starts my attempt at sharing my R journey with the world! I have been learning R off and on now since late 2019,  I have begun to take it much more serious as I work through my Data Analytics class at UCF.  My love for all things numbers and graphs has really blossomed, and I am choosing to share that love with anyone who cares to read.  I will not claim to be the best at R, or any programming for that matter, but these are my attempts.  Each post in this serious will be replicated a graph created in Tableau from the book Tableau for Healthcare.  Todays graph is a simple horizontal bar chart, in transferring to both a new blog site and computer I have unfortunately lost the original bar graph, but trust me the one I created looks just like it. 

# Load Libraries

```{r}
library(tidyr)
library(magrittr)
library(ggplot2)
library(stringr)
library(dplyr)
```

# Import Data

```{r}
ds <- readxl::read_excel(
    path   = "Tableau 10 Training Practice Data.xlsx" 
    ,sheet = "02 - Patient Falls-Single Hosp"
    )
```

# Clean Data Names

```{r}
#should make reusable forumla at later time
names(ds) <- tolower(names(ds))
names(ds) <- str_replace_all(names(ds)," ", "_")

```

# Convert Data to 'Long Form' 

```{r}
ds1 <- ds %>% 
  gather("patient_falls_no_injury_rate" , "patient_falls_with_injury_rate"
         ,key   = "injury" 
         ,value = "rate" ) %>% 
  mutate(injury = (injury == "patient_falls_with_injury_rate"))
         
```

# Graph 5.1

```{r}
b1 <- ds %>% 
  ggplot(mapping = aes(x = reorder(type_of_care,total_patient_falls_rate ) , y = total_patient_falls_rate)) +
  geom_col(fill = "#2b83ba") + 
  coord_flip() +
  scale_y_continuous(breaks = NULL) +
  theme(axis.ticks = element_blank()) +
  labs(title = "Rate of Patient Falls (per 1,000 Pateint Days)\nby Type of Care for FY2017"
      ,x     = NULL
      ,y     = NULL
      ) +
  theme_classic() +
  geom_text(aes(label = format(total_patient_falls_rate, digits = 2)), nudge_y = -.25, color = "white")
  
b1
```
